Fixing Vulnerabilities Hinders Maintainability

### 2019-02-02 paper/ICPC19/icpc19.tex
fixes

This paper is structured as follows: section \ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
(Accessed on \today{})}; section \ref{sec:methodology} describes the
methodology used to answer the research questions; section \ref{sec:results}
presents the results and section \ref{sec:discussion} discusses their
implications; section \ref{sec:threats}, threats to the validity of the study
are enumerated; section \ref{sec:rw} describres the different work and existing
literature in the field of study; and, finally section \ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant and performant
one. Thus, we want to understand how security refactorings impact the
maintainability of software.

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
contains open-source implementations of the SSL and TLS cryptographic
protocols.} to address the fact that hackers could cause a Denial-of-Service
(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.

@@

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
to a potential violation of the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability
of their software application. To answer the following two
research questions, we use a dataset of security refactorings to measure the
impact of the changes in the maintainability of open-source software.

@@

We would expect that the negative impact of programming languages on maintainability
would be more intense due to the buzz of the last years surrounding the idea that
poor design of programming languages and lack of using the best practices lead to
more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}.
However, Figure \ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on
software maintainability. We suspect that these values are the result of projects
contributions policies. 9 out of 10 projects --- the equivalent to 31.8\% of the commits ---
of the top-10 of dataset projects with more contributors have restrict contribution
policies regarding code standards. Some examples are the \texttt{laravel/framework},
\texttt{php/php-src}, \texttt{openssl/openssl} and
\texttt{cakephp/cakephp}. For instance, \texttt{laravel/framework} requires the
\texttt{PSR-2 Standard}\footnote{PSR-2 Standard available at
https://www.php-fig.org/psr/psr-2/ (Accessed on \today{}}).

@@

The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability. Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively. Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added). Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{}}, where any user could login as root with an empty password. Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability. Real efforts have been made in creating and incorporating coding standards and best practices in software production. However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability. Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to validity of this study.
%
\subsection{Construct}
%
The maintainability formula was inferred based on the reports of the Better Code Hub.
Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them. \Sof{Improve it} However, BCH does have a representative benchmark of
closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.

@@

Baseline commits are retrieved randomly by selecting one commit from the same project of a given security refactoring. This approach softens the differences that may result from the
characteristics of each project. However, maintainability still may be affected by the developers experience, coding style and software contribution polices. But maintainability difference is not evaluated at this level.

@@

In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation. Thus, answering these questions to closed commercial applications may be a future work opportunity.

@@

@@

This work presents an empirical study on the impact of 607 security refactorings on
the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
of each project version to calculate maintainability based on a model proposed in
previous work.

Security refactorings significantly affect the software maintainability of open-source
software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
overall maintainability increases (50.16\%). However, there is still a very considerable amout of
cases that harm maintaniability (41.35\%). This study also exhibits evidence that some patterns
need more attention than others. Broken Autentication issues, Denial-of-Service attacks
and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively.

Following the implications of our work, we draw attention for the need of secure security frameworks,
code best practices documentation to produce safe and maintainable code, proper authorization
mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
maintenance and new or improved programming languages to allow developers to easily implement safety in
their software without endangering maintainability.

@@

\noindent
We thank to the Software Improvement Group (SIG) team for all the support and
help in validating our methodology.

>>>>

This paper is structured as follows: section \ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
(Accessed on \today{})} software; section \ref{sec:methodology} describes the
methodology used to answer the research questions; section \ref{sec:results}
presents the results and section \ref{sec:discussion} discusses their
implications; section \ref{sec:threats}, threats to the validity of the study
are enumerated; section \ref{sec:rw} describres the different work and existing
literature in the field of study; and, finally section \ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant and performant
one. Thus, we want to understand how security refactorings impact the
maintainability of software.

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL to address the fact that hackers could
cause a Denial-of-Service (DoS) attack by crafting DTLS
handshake messages to trigger memory allocations corresponding to large length
values. This vulnerability is listed at the Common Vulnerabilities and Exposures
dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at
http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on
\today{})}. It is amongst the vulnerabilities studied in our evaluation.
The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.

@@

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. And, it increases the codebase size
in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

In this study, our concern is to study whether while improving software
security, developers are also inserting a negative impact on the maintainability
of their software which may increase technical debt when choosing the easiest
solution instead of the using the best approach. To answer the following two
research questions, we use a dataset of security refactorings to measure the
impact of the changes in the maintainability of open-source software.

@@

We would expect that the negative impact of programming languages on maintainability 
would be more intense due to the buzz of the last years surrounding the idea that 
poor design of programming languages and lack of using the best practices lead to 
more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. 
However, Figure \ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on 
software maintainability. We suspect that these values are the result of projects 
contributions policies. 9 out of 10 projects --- the equivalent to 31.8\% of the commits --- 
of the top-10 of dataset projects with more contributors have restrict contribution 
policies regarding code standards. Some examples are the \texttt{laravel/framework}, 
\texttt{php/php-src}, \texttt{openssl/openssl} and 
\texttt{cakephp/cakephp}. For instance, \texttt{laravel/framework} requires the 
\texttt{PSR-2 Standard}\footnote{PSR-2 Standard available at 
https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.

@@

The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability. Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively. Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added). Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password. Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability. Real efforts have been made in creating and incorporating coding standards and best practices in software production. However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability. Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to validity of this study.
%
\subsection{Construct}
%
The maintainability formula was inferred based on the reports of the Better Code Hub.
The high amount of different projects and backgrounds may require different maintainablity
standards. However, BCH does have a representative benchmark of closed and open-source 
software projects that is under the calculation of the thresholds used in the computation 
of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.

@@

Baseline commits are retrieved randomly by selecting one commit from the same project of a given security refactoring. This approach softens the differences that may result from the
characteristics of each project. However, maintainability still may be affected by the developers experience, coding style and software contribution polices which it is not evaluated in this study. Furthermore, this evaluation considers that 607 regular commits are enough to alleviate random irregularities in the maintainability differences of the baseline.

@@

In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation. However, BCH thresholds are calculated from a set of both open and closed sotfware. Different programming languages may require different coding practices to address software safety. The dataset
comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population.

@@

\Sof{Talk about language tuning}
% The total volume in a codebase is the volume in lines of code converted to man- months. A man-month is a standard measure of source code volume. It is the amount of source code that one developer with average productivity could write in one month. The advantage of “man-month” is that it allows for comparisons of source code volume between technologies. This is relevant because programming languages have different productivity measures, or “levels of verbosity.” Therefore, a system with multiple programming languages can be converted to an aggregate measure that tells you the approximate effort it would take to rebuild it: the “rebuild value.”

@@

This work presents an empirical study on the impact of 607 security refactorings on
the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
of each project version to calculate maintainability based on a model proposed in 
previous work. Security refactorings significantly affect the software maintainability of open-source 
software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
overall maintainability increases (50.16\%). However, there is still a very considerable amout of
cases that harm maintaniability (41.35\%). This study also exhibits evidence that some patterns
need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following the implications of our work, we draw attention for the need of secure security frameworks,
code best practices documentation to produce safe and maintainable code, proper authorization 
mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
maintenance and new or improved programming languages to allow developers to easily implement safety in
their software without endangering maintainability.

@@

\noindent
We thank to the Software Improvement Group (SIG) team for all the support and
help in validating the methodology.
